Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-02-01 paper.Rnw
Typo fixes from Ainsley

% shortcuts 
\newcommand{\mb}{MAGeT-Brain}
\newcommand{\multiatlas}{multi-atlas}
\newcommand{\ants}{ANTS}
\newcommand{\animal}{ANIMAL}
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T}
\newcommand{\fsl}{FSL FIRST}
\newcommand{\freesurfer}{FreeSurfer}
\newcommand{\maper}{MAPER}
\newcommand{\prsnr}{Pruessner}
\newcommand{\pseg}{\prsnr-protocol manual segmentation}
\newcommand{\psegs}{\prsnr-protocol manual segmentations}
\newcommand{\snt}{SNT}
% experiment #'s 
\newcommand{\expadnixval}{Experiment 1} 
\newcommand{\expfepxval}{Experiment 2} 
\newcommand{\expadni}{Experiment 3}
\newcommand{\expsubfield}{Experiment 4}
\newcommand{\expadnixvalsnt}{Experiment 5}
% experiment descriptions (suitable for titles)
\newcommand{\expadnixvaldesc}{Whole Hippocampus Segmentation Cross-Validation --- Alzheimer's Disease} 
\newcommand{\expfepxvaldesc}{Whole Hippocampus Segmentation Cross-Validation --- First Episode of Psychosis} 
\newcommand{\expadnidesc}{Whole Hippocampus Segmentation Comparision --- ADNI1 Complete 1Yr} 
\newcommand{\expsubfielddesc}{Hippocampal Subfield Segmentation Cross-Validation} 
\newcommand{\expadnixvalsntdesc}{Whole Hippocampus Segmentation Cross-Validation --- Alzheimer's Disease, \snt{} Segmentations}

@@

One methodology that can be used to mitigate these sources of errors involves
the use of multiple manually segmented atlases and probabilistic segmentation
techniques, such as those found in the \freesurfer{} package \citep{Fischl2002}.
\freesurfer{} uses a probabilistic atlas of anatomical and tissue classes along
with spatial constraints for class labels encoded using a Markov random field
model to segment the entire brain.

More recently, many groups have used multiple atlases to improve overall
segmentation accuracy (i.e. \multiatlas{} segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into the definitive segmentation for the target. In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such as cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations. For example \citet{Collins2010} demonstrated that only
14 atlases, selected based on highest similarity between medial temporal lobe
neuroanatomy as evaluated by normalized mutual information \citep{Studholme1999}
from a library of 80 atlases, were required to achieve accurate segmentations of
the hippocampus. Additionally, several methods have been explored for label
fusion, including the STAPLE algorithm (Simultaneous Truth And Performance Level
Estimation; \citet{Warfield2004}) that computes a probabilistic segmentation
using an expectation-maximization framework from an set of competing
segmentations, or others where a subset of segmentations can be estimated using
metrics such as the sum of squared differences in the regions of interest to be
segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library ranging between 30
\citep{Heckemann2006} and 80 \citep{Collins2010} manually segmented atlases.
This strategy has the main drawback of being inflexible as it does not easily
accommodate varying the definition of the hippocampal anatomy (such as the
commonly used heuristic of subdividing the hippocampus into head, body, and tail
\citep{Poppenk2011,Pruessner2000}). Furthermore, none of these methods have
demonstrated sufficient flexibility to accommodate atlases that are somehow
exceptional such as those derived from serial histological data
\citep{Chakravarty2006,Yelnik2007} or high-resolution MRI data that enables
robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}. Due
to the recent availability of the latter, there has been increased interest in
the use of probabilistic methods for the identification of the hippocampal
subfields on standard T1-weighted images. Our group recently demonstrated that
through use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{Chakravarty2013}.
The novelty of this manuscript is the extension of our \multiatlas{} methodology
to the hippocampus using more than a single input atlas, while simultaneously
limiting the number of inputs used during segmentation, and so demonstrating that
reliable application of the atlases' subfield segmentation to an unlabelled
target image is indeed possible using this methodology.

@@

\mb{} ({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates) bootstraps the creation of a large template library given a limited
input atlas library, and then uses the template library in basic \multiatlas
segmentation. Images for the template library are selected from a set of input
target images, either arbitrarily or so as to reflect the neuroanatomy or
demographics of the target set as a whole (for instance, by sampling equally
from cases and controls). The template library images are automatically
labelled by each of the atlases via label propagation. Effectively, basic
\multiatlas{} segmentation is then conducted using the template library to
segment the entire set of target images (including the target images used in the
construction of the template library). Since each template library image has
multiple labels (one from each atlas), the final number of labels to be fused
for each target may be quite large (i.e. \# of atlas $\times$ \# of templates).

@@

\begin{figure}
  \centering
    \includegraphics[width=\textwidth]{figure/MA-MAGeTBrain-Schematic}
  \caption{A schematic illustration of basic \multiatlas{} segmentation and
\mb{} segmentation. In multi-atlas segmentation, manual labels from atlas images
are warped (propagated) into subject space by applying the transformations
estimated from nonlinear image registration. The resulting candidate labels
from all atlas images are then fused to create a final segmentation. In \mb{}
segmentation, a template library is created by sampling (either randomly or
representatively) from the subject images. Atlas labels are propagated to all
template images and then to each subject image (including those used in the template
library). The candidate the labels for a subject are then fused into a final
segmentation. \label{fig:MAGeT}}
\end{figure}

@@

\begin{itemize}
  \item \expadnixval{} investigates \mb{} whole hippocampus
  segmentation of aging and Alzhiemer's diseased subjects over a wide range of
  parameter settings using a Monte Carlo cross-validation design. The results of
  this experiment enable us to choose the parameter settings offering the best
  performance for use in subsequent experiments.

@@

  \item \expadni{} bridges \mb{} with the existing segmentation literature by
  comparing \mb{} whole hippocampus segmentations with those of well-known
  automated methods (\freesurfer, \fsl, \maper, \snt) on the entire
  \adnidataset{} image dataset consistening of
  \Sexpr{length(means.complete$RID)} brain images of subjects diagnosed as
  cognitively normal, having mild cognitive impairment, or Alzheimer's disease.

  \item \expsubfield{} assesses hippocampal subfield segmentation quality in a
  leave-one-out cross-validation on the five high-resolution manually segmented
  Winterburn MR atlases \citep(Winterburn2013).
\end{itemize}

@@

In this experiment we explore the very idea of bootstrapping a template library
for multi-atlas-based segmentation from a small number of input atlases. To do
so, we conduct repeated cross-validations of \mb{} whilst varying the
composition and sizes of the atlas and template libraries used, as well as
varying the registration algorithm and label fusion method. The dataset used in
this experiment is images from the ADNI dataset \citep{Jack2008} along with
whole hippocampus labels manually segmented following the \citep{Pruessner2000}
protocol.

Note, in the Supplementary Materials we have replicated this experiment using
the \snt{} semi-automated segmentations included as part of the the ADNI
dataset.

@@

\paragraph{\adnidataset{} dataset} 
Data used in the preparation of this article were obtained from the Alzheimer's
Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). The ADNI
was launched in 2003 by the National Institute on Aging (NIA), the National
Institute of Biomedical Imaging and Bioengineering (NIBIB), the Food and Drug
Administration (FDA), private pharmaceutical companies and non-profit
organizations, as a \$60 million, 5-year public-private partnership. The primary
goal of ADNI has been to test whether serial magnetic resonance imaging (MRI),
positron emission tomography (PET), other biological markers, and clinical and
neuropsychological assessment can be combined to measure the progression of mild
cognitive impairment (MCI) and early Alzheimer's disease (AD). Determination of
sensitive and specific markers of very early AD progression is intended to aid
researchers and clinicians to develop new treatments and monitor their
effectiveness, as well as lessen the time and cost of clinical trials.

@@

60 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset{}} standardized dataset. 20 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}. Fully manual segmentations of the
left and right whole hippocampi in these images were provided by one author
(JCP) according to the segmentation protocol specified in \citet{Pruessner2000}.

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the "\adnidataset{}" standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.usc.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly pre-processed images
which have been designated to be the "best" after quality control. All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}. Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8^{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel dimensions of $1.25mm \times 1.25mm \times 1.2mm$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<tab:ADNI-xval-demographics, echo=F, dependson='ADNI1-xval-load', results="asis",cache=FALSE>>=
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               jens_atlases$RID, method = "reverse", test=FALSE, overall=TRUE)
caption = 
  "\\textbf{ADNI1 cross-validation subset demographics.} 
  CN     - Cognitively Normal. 
  LMCI   - Late-onset Mild Cognitive Impairment. 
  AD     - Alzheimer's Disease. 
  Hisp   - Hispanic. 
  CDR-SB - Clinical Dementia Rating-Sum of Boxes. 
  ADAS   - Alzheimer's Disease Assessment Scale. 
  MMSE   - Mini-Mental State Examination."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      label="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

  \item[Normalised Mutual Information Weighted Majority Vote]
  This method is similar to cross-correlation weighted voting except that image
  similarity is calculated by the normalised mutual information score over the
  region of interest\citep{Studholme2001}. The {\tt itk\_similarity} utility
  from the EZMinc toolkit\footnote{\url{https://github.com/vfonov/EZminc}} is
  used to calculate the normalised mutual information measure between two images.
\end{description}

@@

We find that for \mb{} segmentations, similarity score increases as atlas and
template library size is increased, although with diminishing returns and an
eventual trend towards a plateau (Figure \ref{fig:ADNI1-xval-k-mean}). A
maximum similarity score of \Sexpr{mb.kappa.best} is found when using 9 atlases,
19 templates, ANTS registration, and majority vote label fusion). The \ants{}
registration method consistently out-performs \animal{} registration over all
variable settings we tested. Additionally, by itself, using a weighted voting
strategy did not significantly improve segmentation agreement, contrary to the
findings of \citet{Aljabar2009} using basic \multiatlas{} segmentation. Given
these findings, in the remainder of our experiments only results using the \ants{}
registration algorithm and majority vote fusion will be shown.

@@

We find similar behaviour with respect to optimal parameter settings and
increased consistency of \mb{} segmentations in the replication of this
experiment (\expadnixvalsnt{}, Supplementary Materials) where a different
hippocampal definition is used (\snt{} labels available with the ADNI datasets).
This strongly suggests that these results are independant of the segmentation
protocol used and are, instead, features of the \mb{} algorithm.

We have omitted results obtained when using an even number of atlases or
templates since with these configurations we found significantly decreased
performance. We believe this is results from an inherent bias in the majority
vote fusion method used (see Discussion).

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 2: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of Psychosis patients                         %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expfepxval{}: \expfepxvaldesc{}}
\label{exp:fep-xval}
To validate that the \mb{} works effectively in the context of other neurological
disorders, in this experiment we replicate the cross-validation done in
\expadnixval{} a dataset of patients having had a single episode of psychosis. We
also compare \mb{} segmentations with those of two well-known automated
segmentation methods, \fsl{} and \freesurfer{}.

@@

Export whole hippocampal manual segmentation of each subject is produced
following a validated segmentation protocol \citep{Pruessner2000}.

@@

\paragraph{Experiment details} 
The same overall design as \expadnixval{} is followed in this experiment: a
Monte Carlo cross-validation (MCCV) is conducted using the pool of 81 first
episode psychosis subject brain images and corresponding \prsnr-protocol manual
segmentations. Five rounds of validation are conducted for each subject, and
each atlas and template library size combination (1-9 atlases, 1-19 templates).
In each round, images and their manual labels are randomly selected from the
pool, and the remaining images are segmented using \mb{} with a random subset of
the unlabelled images also serve as template images. Majority vote fusion, and
the \ants{} registration algorithm are used, as these have shown to behave
favourably in previous experiments.

In addition to the MCCV, we segment the entire first episode psychosis dataset
using \mb{} and two other popular segmentation packages, \fsl{} and
\freesurfer{}. Firstly, \mb{} is used with the five Winterburn atlas images and
labels as atlases, and a randomly selected subset of 19 psychosis dataset images
as templates. Secondly, \mb{} is run using the same template images, but instead
as atlases we use five additional five first episode psychosis subjects and
corresponding manual segmentations (not included above). Finally, \fsl{} and
\freesurfer{} are run with their default settings to segment the entire dataset.
The \fsl{} \verb+run_first_all+ script was used according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}, and
\freesurfer{} was run using \verb+recon-all -all+.

@@

The Pruessner segmentation protocol differs slightly from the Winterburn
protocol those used by \freesurfer{} and \fsl{} in the neuroanatomical features
that are delineated (see \citet{Winterburn2013}, and Table
\ref{tab:subfield-segmentation-protocols} in the Discussion below). This
variation in protocol poses a problem if an overlap measure is used for
evaluation: since different protocols will necessarily produce segmentations
that do not perfectly overlap, the degree of overlap cannot be solely used to
compare segmentation methods using different protocols. In place of an overlap
metric, we assess the degree of (Pearson) correlation in average bilateral
hippocampal volume produced by each method. Additionally, we evaluate the
volume-related fixed and proportional biases in all segmentation methods using
Bland-Altman plots \citep{Bland1986}.

@@

As in \expadnixval, we find that similarity score increases with a greater
number of atlases or templates but quickly plateaus (Figure
\label{fig:FEP-xval}). A maximum similarity score of \Sexpr{snt.mb.kappa.best}
is found when using 9 atlases, 19 templates, ANTS registration, and majority
vote label fusion.

We found a close relationship in average hippocampal volume between the manual
label volumes and \mb{} when using the Winterburn atlases, or manually segmented
FEP subjects as atlases (Figure \label{fig:FEP-vols}). Both sets of volumes are
correlated with Pearson $r > 0.88$. \freesurfer{} and \fsl{} volumes are both
correlated with manual volumes at Pearson $r > 0.7$.

>>>>

% shortcuts 
\newcommand{\mb}{MAGeT-Brain}
\newcommand{\multiatlas}{multi-atlas}
\newcommand{\ants}{ANTS}
\newcommand{\animal}{ANIMAL}
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T}
\newcommand{\fsl}{FSL FIRST}
\newcommand{\freesurfer}{FreeSurfer}
\newcommand{\maper}{MAPER}
\newcommand{\prsnr}{Pruessner}
\newcommand{\pseg}{\prsnr-protocol manual segmentation}
\newcommand{\psegs}{\prsnr-protocol manual segmentations}
\newcommand{\snt}{SNT}
% experiment #'s 
\newcommand{\expadnixval}{Experiment 1} 
\newcommand{\expfepxval}{Experiment 2} 
\newcommand{\expadni}{Experiment 3}
\newcommand{\expsubfield}{Experiment 4}
\newcommand{\expadnixvalsnt}{Experiment 5}
% experiment descriptions (suitable for titles)
\newcommand{\expadnixvaldesc}{Whole Hippocampus Segmentation Cross-Validation --- Alzheimer's Disease} 
\newcommand{\expfepxvaldesc}{Whole Hippocampus Segmentation Cross-Validation --- First Episode of Psychosis} 
\newcommand{\expadnidesc}{Whole Hippocampus Segmentation Comparison --- ADNI1 Complete 1Yr} 
\newcommand{\expsubfielddesc}{Hippocampal Subfield Segmentation Cross-Validation} 
\newcommand{\expadnixvalsntdesc}{Whole Hippocampus Segmentation Cross-Validation --- Alzheimer's Disease, \snt{} Segmentations}

@@

One methodology that can be used to mitigate these sources of error involves
the use of multiple manually segmented atlases and probabilistic segmentation
techniques, such as those found in the \freesurfer{} package \citep{Fischl2002}.
\freesurfer{} uses a probabilistic atlas of anatomical and tissue classes along
with spatial constraints for class labels encoded using a Markov random field
model to segment the entire brain.

More recently, many groups have used multiple atlases to improve overall
segmentation accuracy (i.e. \multiatlas{} segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into the definitive segmentation for the target. In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such as cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations. For example \citet{Collins2010} demonstrated that only
14 atlases, selected based on highest similarity between medial temporal lobe
neuroanatomy as evaluated by normalized mutual information \citep{Studholme1999}
from a library of 80 atlases, were required to achieve accurate segmentations of
the hippocampus. Also, several methods have been explored for label
fusion. For example, the STAPLE algorithm (Simultaneous Truth And Performance
Level Estimation; \citet{Warfield2004}) uses an expectation-maximization
framework to compute a probabilistic segmentation from a set of competing
regmentations, or the work of \citet{Coupe2011} who show that a subset of
segmentations can be estimated using metrics, such as the sum of squared
differences in the regions of interest to be segmented.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library ranging between 30
\citep{Heckemann2006} and 80 \citep{Collins2010} manually segmented atlases.
This strategy has the main drawback of being inflexible as it does not easily
accommodate varying the definition of the hippocampal anatomy (such as the
commonly used heuristic of subdividing the hippocampus into head, body, and tail
\citep{Poppenk2011,Pruessner2000}). Furthermore, none of these methods have
demonstrated sufficient flexibility to accommodate atlases that are somehow
exceptional such as those derived from serial histological data
\citep{Chakravarty2006,Yelnik2007} or high-resolution MRI data that enables
robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}. Due
to the recent availability of the latter, there has been increased interest in
the use of probabilistic methods for the identification of the hippocampal
subfields on standard T1-weighted images. Our group recently demonstrated that
through use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{Chakravarty2013}.
The novelty of this manuscript is the extension of our \multiatlas{} methodology
to the segmentation of hippocampus. Additionally, in this paper we rigorously explore the
effects of using multiple input atlases, of varying the size of the template
library constructed, and registration and label fusion methods. As a result, we
aim to demonstrate that it is indeed possible to reliably apply the segmentation
represented in a very small set of segmented input atlases to an unlabelled
target image set.

@@

\mb{} ({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates) bootstraps the creation of a large template library given a limited
input atlas library, and then uses the template library in basic \multiatlas{}
segmentation. Images for the template library are selected from a set of input
target images, either arbitrarily or so as to reflect the neuroanatomy or
demographics of the target set as a whole (for instance, by sampling equally
from cases and controls). The template library images are automatically
labelled by each of the atlases via label propagation. Effectively, basic
\multiatlas{} segmentation is then conducted using the template library to
segment the entire set of target images (including the target images used in the
construction of the template library). Since each template library image has
multiple labels (one from each atlas), the final number of labels to be fused
for each target may be quite large (i.e. \# of atlas $\times$ \# of templates).

@@

\begin{figure}
  \centering
    \includegraphics[width=\textwidth]{figure/MA-MAGeTBrain-Schematic}
  \caption{A schematic illustration of basic \multiatlas{} segmentation and
\mb{} segmentation. In multi-atlas segmentation, manual labels from atlas images
are warped (propagated) into subject space by applying the transformations
estimated from nonlinear image registration. The resulting candidate labels
from all atlas images are then fused to create a final segmentation. In \mb{}
segmentation, a template library is created by sampling (either randomly or
representatively) from the subject images. Atlas labels are propagated to all
template images and then to each subject image (including those used in the template
library). The candidate labels for a subject are then fused into a final
segmentation. \label{fig:MAGeT}}
\end{figure}

@@

\begin{itemize}
  \item \expadnixval{} investigates \mb{} whole hippocampus
  segmentation of aging and Alzheimer's diseased subjects over a wide range of
  parameter settings using a Monte Carlo cross-validation design. The results of
  this experiment enable us to choose the parameter settings offering the best
  performance for use in subsequent experiments.

@@

  \item \expadni{} bridges \mb{} with the existing segmentation literature by
  comparing \mb{} whole hippocampus segmentations with those of well-known
  automated methods (\freesurfer, \fsl, \maper, \snt) on the entire
  \adnidataset{} image dataset consisting of \Sexpr{length(means.complete$RID)}
  brain images of subjects diagnosed as cognitively normal, having mild cognitive
  impairment, or Alzheimer's disease.

  \item \expsubfield{} assesses hippocampal subfield segmentation quality in a
  leave-one-out cross-validation on the five high-resolution manually segmented
  Winterburn MR atlases \citep{Winterburn2013}.
\end{itemize}

@@

In this experiment we explore the very idea of bootstrapping a template library
for multi-atlas-based segmentation from a small number of input atlases. To do
so, we conduct repeated cross-validations of \mb{} whilst varying the
composition and sizes of the atlas and template libraries used, as well as
varying the registration algorithm and label fusion method. The dataset used in
this experiment is images from the ADNI dataset \citep{Jack2008} along with
whole hippocampus labels manually segmented following the \prsnr-protocol
\citep{Pruessner2000}.

Note, in the Supplementary Materials we have replicated this experiment using
the \snt{} semi-automated segmentations included as part of the ADNI
dataset.

@@

\paragraph{\adnidataset{} dataset} 
\label{adnidataset}
Data used in the preparation of this article were obtained from the Alzheimer's
Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). The ADNI
was launched in 2003 by the National Institute on Aging (NIA), the National
Institute of Biomedical Imaging and Bioengineering (NIBIB), the Food and Drug
Administration (FDA), private pharmaceutical companies and non-profit
organizations, as a \$60 million, 5-year public-private partnership. The primary
goal of ADNI has been to test whether serial magnetic resonance imaging (MRI),
positron emission tomography (PET), other biological markers, and clinical and
neuropsychological assessment can be combined to measure the progression of mild
cognitive impairment (MCI) and early Alzheimer's disease (AD). Determination of
sensitive and specific markers of very early AD progression is intended to aid
researchers and clinicians to develop new treatments and monitor their
effectiveness, as well as lessen the time and cost of clinical trials.

@@

Sixty 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset{}} standardized dataset. Twenty subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}. Fully manual segmentations of the
left and right whole hippocampi in these images were provided by one author
(JCP) according to the segmentation protocol specified in \citet{Pruessner2000}.

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the \adnidataset{} standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.usc.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly pre-processed images
which have been designated to be the ``best'' after quality control. All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citet{Jack2008}. Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8^{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding voxel dimensions of $1.25mm \times 1.25mm \times 1.2mm$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<tab:ADNI-xval-demographics, echo=F, dependson='ADNI1-xval-load', results="asis",cache=FALSE>>=
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               jens_atlases$RID, method = "reverse", test=FALSE, overall=TRUE)
caption = 
  "\\textbf{ADNI1 cross-validation subset demographics.} 
  CN     - Cognitively Normal. 
  LMCI   - Late-onset Mild Cognitive Impairment. 
  AD     - Alzheimer's Disease. 
  CDR-SB - Clinical Dementia Rating-Sum of Boxes. 
  ADAS   - Alzheimer's Disease Assessment Scale. 
  MMSE   - Mini-Mental State Examination."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      label="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

  \item[Normalised Mutual Information Weighted Majority Vote]
  This method is similar to cross-correlation weighted voting except that image
  similarity is calculated by the normalised mutual information score over the
  region of interest \citep{Studholme2001}. The {\tt itk\_similarity} utility
  from the EZMinc toolkit\footnote{\url{https://github.com/vfonov/EZminc}} is
  used to calculate the normalised mutual information measure between two images.
\end{description}

@@

We find that for \mb{} segmentations, similarity score increases as atlas and
template library size is increased, although with diminishing returns and an
eventual trend towards a plateau (Figure \ref{fig:ADNI1-xval-k-mean}). A
maximum similarity score of \Sexpr{mb.kappa.best} is found when using 9 atlases,
19 templates, ANTS registration, and majority vote label fusion. The \ants{}
registration method consistently out-performs \animal{} registration over all
variable settings we tested. Additionally, by itself, using a weighted voting
strategy did not significantly improve segmentation agreement, contrary to the
findings of \citet{Aljabar2009} using basic \multiatlas{} segmentation. Given
these findings, in the remainder of our experiments only results using the \ants{}
registration algorithm and majority vote fusion will be shown.

@@

We find similar behaviour with respect to optimal parameter settings and
increased consistency of \mb{} segmentations in the replication of this
experiment (\expadnixvalsnt{}, Supplementary Materials) where a different
hippocampal definition is used (\snt{} labels available with the ADNI datasets).
This strongly suggests that these results are independent of the segmentation
protocol used and are, instead, features of the \mb{} algorithm.

We have omitted results obtained when using an even number of atlases or
templates since with these configurations we found significantly decreased
performance. We believe this results from an inherent bias in the majority vote
fusion method used (see Discussion).

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 2: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of Psychosis patients                         %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expfepxval{}: \expfepxvaldesc{}}
\label{exp:fep-xval}
To validate that the \mb{} works effectively in the context of other
neurological disorders, in this experiment we replicate the cross-validation
done in \expadnixval{} with a dataset of patients having had a single episode of
psychosis. We also compare \mb{} segmentations with those of two well-known
automated segmentation methods, \fsl{} and \freesurfer{}.

@@

Expert whole hippocampal manual segmentation of each subject is produced
following a validated segmentation protocol \citep{Pruessner2000}.

@@

\paragraph{Experiment details} 
The same overall design as \expadnixval{} is followed in this experiment: a
Monte Carlo cross-validation (MCCV) is conducted using the pool of 81 first
episode psychosis subject brain images and corresponding \prsnr-protocol manual
segmentations. Five rounds of validation are conducted for each subject, and
each atlas and template library size combination (1-9 atlases, 1-19 templates).
In each round, images and their manual labels are randomly selected from the
pool, and the remaining images are segmented using \mb{} with a random subset of
the unlabelled images also serving as template images. Majority vote fusion, and
the \ants{} registration algorithm are used, as these have shown to behave
favourably in previous experiments.

In addition to the MCCV, we segment the entire first episode psychosis dataset
using \mb{} and two other popular segmentation packages, \fsl{} and
\freesurfer{}. Firstly, \mb{} is used with the five Winterburn atlas images and
labels as atlases, and a randomly selected subset of 19 psychosis dataset images
as templates. Secondly, \mb{} is run using the same template images, but instead
as atlases we use five additional first episode psychosis subjects and
corresponding manual segmentations (not included above). Finally, \fsl{} and
\freesurfer{} are run with their default settings to segment the entire dataset.
The \fsl{} \verb+run_first_all+ script was used according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}, and
\freesurfer{} was run using \verb+recon-all -all+.

@@

The \prsnr{} segmentation protocol differs slightly from the Winterburn
protocol, and those used by \freesurfer{} and \fsl{}, in the inclusion of
neuroanatomical features and the manner they are delineated (see
\citet{Winterburn2013}, and Table \ref{tab:subfield-segmentation-protocols} in
the Discussion below). This variation in protocol poses a problem if an overlap
measure is used for evaluation: since different protocols will necessarily
produce segmentations that do not perfectly overlap, the degree of overlap
cannot be solely used to compare segmentation methods using different protocols.
In place of an overlap metric, we assess the degree of (Pearson) correlation in
average bilateral hippocampal volume produced by each method. Additionally, we
evaluate the volume-related fixed and proportional biases in all segmentation
methods using Bland-Altman plots \citep{Bland1986}.

@@

As in \expadnixval, we find that similarity score increases with a greater
number of atlases or templates but quickly plateaus (Figure \ref{fig:FEP-xval}).
A maximum similarity score of \Sexpr{snt.mb.kappa.best} is found when using 9
atlases, 19 templates, ANTS registration, and majority vote label fusion.

We found a close relationship in average hippocampal volume between the manual
label volumes and \mb{} when using the Winterburn atlases, or manually segmented
FEP subjects as atlases (Figure \ref{fig:FEP-vols}). Both sets of volumes are
correlated with Pearson $r > 0.88$. \freesurfer{} and \fsl{} volumes are both
correlated with manual volumes at Pearson $r > 0.7$.

As \citet{Bland1986} noted, high correlation amongst measures of the same
quantity does not necessarily imply agreement (as correlation can be driven by a
large range in true values, for instance). Figure \ref{fig:FEP-Bland-Altman}
shows Bland-Altman plots illustrating the level of agreement of each method with
manual volumes. All methods show an obvious proportional bias: \freesurfer{}
and \fsl{} markedly underestimate smaller hippocampi and over-estimate large
hippocampi (the limits of agreement are between $-2482^3mm$ and $-784^3mm$, and
between $-1653^3mm$ and $79^3mm$, respectively), whereas both \mb{} methods show
a much less exaggerated, but conservative bias (limits of agreement between
$-67^3mm$ and $766^3mm$ when using FEP atlases, and between $-333^3mm$ and
$504^3mm$ when using Winterburn atlases). On average, \freesurfer{} and \fsl{}
overestimate hippocampal volume by about $1600 mm^3$ and $800 mm^3$,
respectively. In contrast, on average \mb{} underestimates volumes by about $300
mm^3$ when using FEP atlases, and by about $80mm^3$ when using Winterburn
atlases.
